Agile Formality: A ``Mole'' of Software Engineering Practices

### 2010-07-05 papers/AgileFormality/paper.tex
Renaming directory to more closely reflect contents of paper.

\title{Agile Formality: A Mole of Software Engineering Practices}

@@

  Members of the agile programming and formal methods communities do
  not always see eye-to-eye.  These two communities often do not talk
  to or learn from each other.  Only recently, as highlighted by the
  September 2009 issue of IEEE Software, the IFIP workshop on
  balancing agility and formalism in software engineering, and the
  first edition of the first international workshop for formal methods
  and agile methods, have ideas from the two communities begun to be
  synthesized.  While the problem-solving approaches and psychological
  attitudes of members of the two communities differ widely, we
  exploit this clash of viewpoints, synthesizing a new development
  processes that actually blends, rather than mashes together, best
  practices from the two worlds.  This paper summarizes our process
  and a supporting complex case study, showing that it is not only
  possible, but tasty, to combine the ``chili pepper'' of formal
  methods and the ``chocolate'' of agile programming, thus producing a
  tasty ``mole'' (as in the highly-spices Mexican sauce) of software
  engineering practices.

@@

% agile and formal difficult to blend: mind attitudes

Agile and formal development methodologies
usually do not blend together well.  This is because of several
reasons, the most important of which is often characterized as a
radical difference in psychological attitudes about software
development.

Formal methodologists often favour an in-depth, think-first approach,
where the problem is understood, formalized, and solved; usually, but
not always, adopting a waterfall-style of development.  Once a
formalization is developed, development and verification proceeds, in
an automated and interactive fashion.  Consequently, projects that use
formal methodologies (FMs) typically focus on critical systems with
fixed requirements and somewhat flexible deadlines (``We'll ship it
when it is right.'')---systems in which if something goes wrong,
someone dies or, directly or indirectly, looses a very large amount of
money.

Agile methodologists favour an highly incremental and iterative
approach, specifically tailored to cope with changing requirements and
precise deadlines.  In projects that use agile methodologies (AMs), it is
often the case that the problem is only partially understood.  Thus,
one focuses on the aspects of a system that must be implemented in the
current development iteration.  The solution initially developed is
usually not optimal at all, but as it is refined through continuous
code refactorings, its quality and validity improve over time.  Test
suites are used for several purposes, the most important of which are
system verification, system documentation, and requirement
specification, and they are strictly handmade.  AMs are typically used
in development settings where changing requirements and rapid delivery
of the product are paramount.

% problems would benefit form agile and formal mix, characterize the problem

The two worlds seem irreconcilable.  Nevertheless, there exist
problems that would greatly benefit from both~\cite{Black2009}.  The
problem class that in which we have particular interest have
\emph{unstable requirements}, are \emph{constrained by deadlines that
  cannot be postponed}, but also \emph{have subsystems that must be
  formally specified and verified}.

% terminology

@@

In this paper the IEEE 610~/cite{Jay1990} standard terminology for validation and
verification is used.  Paraphrasing the standard
slightly, \emph{verification} is the process of evaluating software to
determine whether the products of a given development phase satisfy
the conditions imposed at the start of that phase, according to the
requirements and design specifications (``you build the system in the
right way''); validation is the process of evaluating software during
or at the end of the development process to determine whether it
satisfies specified requirements ensuring it meets the user's needs
(``you build the right system'').

Within the formal methods community, verification means something much
stronger, and validation something quite different, than in the IEEE
610 standard.  Loosely, within the FM community, \emph{verification}
means the use of formal methods to formally, statically (and usually
modularly) verify that a system conforms to its specification (under
all possible execution scenarios).  On the other hand,
\emph{validation} means the use of traditional software engineering
practices to check, statically or dynamically, that the system
conforms to its specification (whether in the form of tests,
requirements, etc.) by hand or via execution under some execution
scenarios.

As we come from both the AM and FM communities, we respect and
acknowledge both perspectives and will attempt to be absolutely clear
in our use of these terms in the following.

Additionally, in the following when we informally say ``...development
of a system,'' we more precisely mean ``...analysis, design,
development, verification (both testing and formal verification), and
validation of a system.''

@@

Proposing a new process with no supporting evidence is an
uninteresting proposition.  A complex case study is mandatory to test
bench a development process.  Our case study focuses on the
development of a device driver to control and communicate with an
embedded custom circuit board equipped with more than a dozen sensors.
The communication protocol with the board is asynchronous and packet
based and the board is novel and custom-designed.  The protocol is
focused on controlling the board and reporting sensor measurements and
board status.

This software driver has been developed using our process while
statisfying seemingly contradictory requirements: (1)~the protocol
implementation must be thoroughly verified; (2)~parts of the protocol
must be formally specified; (3)~a board simulator must be designed,
implemented, and verified to match the specifications of, and actual
behavior of, the physical circuit board; (4)~if board components
change during development, the associated formal specifications and
the simulator whose behavior reflects such changes must be updated
accordingly; (5)~strict software delivery deadlines.  Consequently,
these process requirements constitute an environment that benefits
from both an agile and a formal approaches to software development.

Over the course of the past 18 months, an novel development process
has been refined to cope with the contrasting requirements of the case
study, blending agile and formal practices.  In the following we
summarize this development process and report on its successful
application to the case study, yielding several promising results.

@@

% what is needed, high-level
To enable the integration of the two worlds, the highly iterative and incremental approach found in most of the agile methodologies must be maintained, and formal verification  methodologies, traditionally used in a waterfall development process (formally specify the system, implement the system, verify the implementation against the specification), must be adapted to a highly iterative one.

% what must be considered: TDD
The most common software verification practices in agile methodologies must be considered, when trying to integrate formal and agile methodologies, since they are a fundamental part of the development process. 
The most significant verification practice is Test Driven Development (TDD)~\cite{Beck2003}, TDD is a software development technique, originally defined in Extreme Programming (XP)~\cite{Beck2004} methodology, characterized by writing unit tests before writing the code.
Actually, the general technique, consisting in writing the tests before the code, is used at high abstraction levels as well, since user stories (the requirements) are translated into test cases (the functional tests) before implementation.
The test driven approach is a cornerstone of XP, but it is so popular that have been adopted in many other agile methodologies as well.
It is also considered a good software development technique when used on its own, regardless of the enclosing development process.

% test driven kinds
The test driven approach relies on writing the tests before the actual implementation. 
It is appliable in a very small scale (Test Driven Development) yielding to activity cycles as short as a few minutes.
It ia appliable in a large scale as well (requirements immediately translated into functional tests), yeilding to activity cycles as long as a full delivery iteration (usually no longer than a couple of months).

% previous related work
Some tentative attempts to reconcile formal and agile methodologies have been developed.
The recurrent approach of the attempts consists in removing the handmade test suites, substituting them with different kinds of automated verification based on the system formal specification. 
The automated verification can be as simple as enabling run time verification of assertions or automatically generating a test suite based on the system properties, or as complex as actually proving the whole system behaviour through static model and code checkers or theorem provers.

% a quick analysis of the related works
Eleftherakis and Cowling in~\cite{Eleftherakis2003} propose XFun, an iterative and incremental process to generate X-Machines models, an extension of Finite State Machines. 
The development process impose a complete specification of the system, and its verification is done exclusively using the tools provided by the X-Machines formal methodology. 
The development process proposed in~\cite{Herranz2003b} by Herranz and Moreno-Navarro is quite similar; they propose an iterative process to model a system using SLAM-SL, an object-oriented specification language, and its tools. 
Some XP practices are fully adopted or considered compatible, as pair-programming, iterative and incremental development, system refactoring; while tests are completely replaced by the verification tools provided by the SLAM suite.
A different formal methodology, but same approach, is found also in~\cite{Suhaib2005} where XFM methodology is explained.

% previous work: a common approach
As already mentioned, in all these attempts, the handmade test suites are automatically generated using the system formal specification, or completely replaced by theorem provers or static checkers. 
The requisite to enable these approaches is, at least, a specification of the entire system under development to enable simple verification techniques as run time verification (obtained by assertions and design by contract~\cite{Meyer1997} preconditions, postconditions and invariants) and test suites automatic generation~\cite{Cheon2002,Cheon2004,Cheon2005}.
At most, a complete and sound system specification, usually supported by an appropriate formal language (to specify model and properties) and a constrained programming language, used to actually prove properties about the system under development~\cite{CatanoHuisman02,DetlefsNelsonSaxe2005,KiniryCok04}.

% previous work: why is not satisfying
The approach is simple: the verification, documentation and requirement specification purposes of the tests are moved on to the formal specifications, and a complete formal specification of the system under development is needed to apply these methodologies.
The problem is that this requirement is hard to fulfil, since the effort required to write a complete formal specification of the system, in most real world complex cases, is usually greater than writing a suite of tests~/cite{Glass2002}.
If a complete formal specification is not built, all the previous methodologies share a common problem: the parts not formally specified are not verified at all.
The parts not formally specified can obviously be verified with the traditional methods, but the loops between development artifacts and related activities (code, tests, design, refactoring) are not detailed nor enforced.
The loops and relations between the different development artifacts are the inner engine of many agile methodologies, they impose an iterative and incremental pace to the development process.

% what we want to achieve
The objectives we want to achieve are the following ones: freedom to formally specify only parts of the system, in order to cope with constrained resources and unstable requirements; to draft an highly iterative and incremental development process that blends formal and agile practices.

% how we want to achive it
The main specific problem, by now never addressed, is how to connect development artifacts of the two worlds together: we will define activity cycles, similar to what is found in agile methodologies (especially in XP), to solve this problem in an highly iterative development process.
We will show how tests can drive the formal specification, how the formal specifications can drive tests and code development, how handmade and automated tests can coexist and support each other, how the unspecified parts of the system can be incrementally specified.
That is, we will show how to blend, and not compose, formal and agile verification practices: a Mole of verification practices\footnote{A mole (\emph{'m\=ol\=a}) is a highly spiced Mexican sauce made chiefly from chili peppers and chocolate, served with meat.}

@@

Mole practices are applied in the context of a real world case study
that matches the problem requirements previously detailed: unstable
requirements and development artifacts that need to be formally
specified and verified.  The case study fits these needs.

@@

% SenseTile introduction
The \textbf{UCD CASL SenseTile System} is a large-scale,
general-purpose sensor system developed at the University College
Dublin in Dublin, Ireland.  Our facility provides a scalable and
extensible infrastructure for performing in-depth investigation into
both the specific and general issues in large-scale sensor networking.
This system integrates a sensor platform, a datastore, and a compute
farm.  The sensor platform is a custom-designed but inexpensive sensor
platform (called the \ST) paired with general-purpose small-scale
compute nodes, including everything from low-power PDAs to powerful
touchscreen-based portable computers.  Besides containing over a dozen
sensors packaged on the \ST itself, the board is expandable as well,
as new sensors are added to it easily.

% case study: the sensor board development tasks
The case study is focused in building the sensor board and its
software driver concurrently.  Because of hard time constraints and
the initial unavailability of the custom board, we must progress
concurrently with all the development tasks, including:
(1)~specification of the communication protocol; (2)~specification of
the physical sensor board; (3)~development and fabrication of the
physical board; (4)~development of the embedded software for the
board; (5)~development of the communication protocol software driver
the the board; and (6)~development of a software simulator of the
board.

% what we have to do, what we outsourced, and the dependencies between tasks
The development of the physical board, along with its embedded
software (development tasks~2, 3, 4), is carried out by an third party
under our guidance, thus it is not directly taken into account here.
The specification of the communication protocol (development task~1)
is a joint effort between ourselves and the third party, while the
remaining tasks~(5, 6) are performed in isolation.  Dependencies are
straightforward: the specification of the communication protocol~(1)
is the most stable element, but it still depends on the sensor
board~(2, 3, 4): large changes in the latter affect the former.
Additionally, the driver and the simulator depends directly on the
communication protocol.

% communication with external manufacturer
Communication and synchronization with the external manufacturer is
frequent but not optimal.  The main synchronization artifact is the
protocol specification, which remains stable at high-level, but is
changed frequently in the low-level details.  Since domain expertise
is very different between the two teams, a common domain language is
difficult to find and to agree upon.  Formal methodologies and
programming platforms also differ.  Consequently, more for social and
psychological reasons than technical ones, the informal specification
of the communication protocol is the most reliable source of
information.

@@

The hardware-software formal co-design methodologies usually have
several common development artifacts~\cite{Slomka2000, Hoffman2001}
including a high-level specification of the system, a translation
(refinement) of the high-level specification to low-level ones
(hardware and software counterparts), the possibility to generate
software code from the low-level specifications, a hardware simulator
capable of simulating an hardware component based on hardware
specification, the hardware component developed.

When considering the development of the device and of its software
driver as separate entities that possibly have to be developed
concurrently, the existing approaches are similar to the ones seen in
the case of hardware-software co-design~\cite{Valderrama1995,
  Siegmund2002, Ryzhyk2009}, all of which focus on a specification
that aims to be as complete.

% complete specification is not feasible
A complete specification is not feasible nor convenient in our case.
The protocol specification must be enhanced incrementally and its
complete specification cannot be provided.  This makes the problem an
ideal candidate to test the our ``mole'' of practices.

% formal methodology constraints
To perform this case study, an appropriate formal methodology must be
chosen.  It must be able to support a specification that is built
incrementally.  Consequently, formal methods that demand a complete
specification must be avoided.  Additionally,
to best synthesize with agile methods, the formal language must be
able to automatically generate tests.  Finally, the formal method must
be complemented by tools that support the verification of system
properties both at design-time (via static analysis, for full
functional verification of critical subsystems) and at run time (for
dynamic validation for integration with TDD).

% the chosen formal methodology
Driven by these requirements and our expertise, the Java programming
language and platform were chosen for the implementation, the Java
Modeling Language (JML)~\cite{LeavensBakerRuby-Prelim} and Business
Object Notation (BON)~\cite{WaldenNerson95} are used to write formal
specifications, and the BONc~\cite{BONc}, Beetlz~\cite{Beetlz},
JML2~\cite{JML2,JML2common}, and Extended Static Checker for Java
(ESC/Java2)~\cite{ESCJava2,KiniryCok04SmallCite} tool suites are used
as supporting tools~\cite{BurdyEtal05-STTT}.  The JML2 tool suite
includes runtime assertion checker~\cite{Cheon-Leavens02b} and a unit
test generator~\cite{Cheon2002}.  The BONc tool supports validity
checking of high-level requirements and architecture specifications,
Beetlz supports reasoning about refinement between BON and JML
specifications, and ESC/Java2 supports automatic, modular, static
reasoning about the correctness of an implementation against its
formal specification.  All four tool suites do not require a complete
specification to be used effectively.

\subsection{\todo{paper structure title}}

\todo{explain paper structure}

% ======================================================================
\section{Agile Methodologies}
\label{sec:agile_methodologies}

% abstraction in test suites
High level test suites, composed of functional tests, describe requirements and bind together requirements documentationand system behaviour: an informal requirement is supported by a set of functional tests. 
The associated development cycle is one iteration long. 
Low level test suites, composed of unit tests, describe the modules behaviour and bind together code documentation, code and the contract imposed on the module: unit tests represent the contract and the documentation. 
The associated development cycle is less than one day (as short as a few minutes).

% ======================================================================
\section{Formal Specifications With JML}
\label{sec:formal_specifications_using_jml}

% JML introduction
The Java Modeling Language (JML) is the de facto standard
specification language for describing the behavioural properties of
Java programs.  JML is a rich behavioural interface specification
language (BISL) and thus focuses on the modular specification of
classes and objects in which inheritance is used solely for
behavioural subtyping~\cite{LeavensBakerRuby99}.  JML includes
standard specification constructs like assumptions, assertions,
axioms, and pre- and postconditions, framing clauses, and invariants.
It also includes a rich model-based specification layer that includes
mathematical models, data refinements~\cite{}, datagroups~\cite{}, and
many other advanced object-oriented specification
features~\cite{chalin06:_beyon_asser}.  Many tools, ranging from
compilers to static checkers to full-functional verification systems
support the JML language~\cite{BurdyEtal05-STTT}.

@@

@@

% composition of the two worlds
Both Test Driven Development (TDD) and classical Formal specification Driven Development (FDD) guarantee consistency on all the involved artifacts. 
Both TDD and FDD support iterative development.
The development cycle adopted in TDD is shown in Figure \ref{fig:tdd_diagram} and the one adopted in classic FDD is shown in Figure \ref{fig:fdd_diagram}.
Unit tests, in TDD, are substituted, in FDD, by formal specifications supported by the verification environment, in our case the verification environment is the generated unit test suite.
Formal specifications replace unit tests: they specify, document and verify the system under development.

\begin{figure}[ht]
  \begin{minipage}[b]{0.49\linewidth}
    \centering
    \includegraphics[scale=0.5]{diagrams/TDD_classic}
    \caption{TDD development cycle.}
    \label{fig:tdd_diagram}
  \end{minipage}
  \hspace{0.02\linewidth}
  \begin{minipage}[b]{0.49\linewidth}
    \centering
    \includegraphics[scale=0.5]{diagrams/FDD_classic}
    \caption{FDD classic development cycle.}
    \label{fig:fdd_diagram}
  \end{minipage}
\end{figure}

% composition is not enough
Is this enough to deal with formal and informal artifacts?
FDD is used for the formal specified parts of the system, whether TDD is used for the remaining ones: a ``composition'' of formal and agile verification practices.
Yet, it is not enough.
There are several problem that still needs to be addressed.
No communication between the two worlds is permitted, it is not defined how to formally specify a module after it was implemented and verified in a non formal way.
A module only partially specified is a sort of hybrid that must be treated accordingly.
Finally, in the FDD cycle shown in Figure \ref{fig:fdd_diagram}, the implicit assumption is that it is always possible to identify the correct formal specifications of the system, but this is not always straighforward.
Identify the correct formal specification is a difficult task, it has to be supported and verified by the development process.

% FDD with handmade tests
A partially specified module cannot be verified properly only through automatically generated tests.
Handmade tests have to be used together with generated ones.
Handmade tests verify complex behaviours that are not formally specified, or complex behaviours that are formally specified but cannot be replicated by the formal verification tools, because of the limits of the verification tools themselves. 
The constraints imposed by formal verification tools~\cite{Glass2002} must be taken into account when deciding whether supporting the code with handmade tests, and verification tools based on automatically generated tests are not an exception.
The development cycle able to blend formal specifications, handmade tests and code development is shown in Figure \ref{fig:fdd_hand_made_tests_diagram}: the formal specification drives the handmade tests, both handmade tests and the formal specification drive the code development.

>>>>

\title{Agile Formality: A ``Mole'' of Software Engineering Practices}

@@

Members of the agile programming and formal methods communities do not always see eye-to-eye.  
These two communities often do not talk to or learn from each other.  
Only recently, as highlighted by the September 2009 issue of IEEE Software, the IFIP workshop on balancing agility and formalism in software engineering, and the first edition of the international workshop for formal methods and agile methods, ideas from the two communities begun to be synthesized.  
While the problem-solving approaches and psychological attitudes of members of the two communities differ widely, we exploit this clash of viewpoints, exploiting a new development processes that actually blends, rather than mashes together, best practices from the two worlds.  
This paper summarizes our process and a supporting complex case study, showing that it is not only possible, but tasty, to combine the ``chili pepper'' of formal methods and the ``chocolate'' of agile programming, thus producing a tasty ``Mole'' (as in the highly-spices Mexican sauce) of software engineering practices.

@@

% agile and formal difficult to blend: mind attitudes
Agile and formal development methodologies usually do not blend together well.  
This is because of several reasons, the most important of which is often characterized as a radical difference in psychological attitudes about software development.

% FM
Formal methodologists often favour an in-depth, think-first approach, where the problem is understood, formalized, and solved; usually, but not always, adopting a waterfall-style of development.  
Once a formalization is developed, development and verification proceed, in an automated and interactive fashion.  
Consequently, projects that use formal methodologies (FMs) typically focus on critical systems with fixed requirements and somewhat flexible deadlines (``We'll ship it when it is right.'').

% AM
Agile methodologists favour an highly incremental and iterative approach, specifically tailored to cope with changing requirements and precise deadlines.  In projects that use agile methodologies (AMs), it is often the case that the problem is only partially understood.  
Thus, one focuses on the aspects of a system that must be implemented in the current development iteration.  
The solution initially developed is usually not optimal at all, but as it is refined through continuous code refactorings, its quality and validity improve over time.  
Test suites are used for several purposes, the most important of which are system verification, system documentation, and requirement specification, and they are strictly handmade.  
AMs are typically used in development settings where changing requirements and rapid delivery of the product are paramount.

% problems would benefit form agile and formal mix, characterize the problem
The two worlds seem irreconcilable.  
Nevertheless, problems exist that would greatly benefit from both of them~\cite{Black2009}.  
The problem class in which we have particular interest have \emph{unstable requirements}, are \emph{constrained by deadlines that cannot be postponed}, but also have \emph{subsystems that must be formally specified and verified}.

@@

% terminology
In this paper the IEEE 610~/cite{Jay1990} standard terminology for validation and verification is used.  
Paraphrasing the standard slightly, \emph{verification} is the process of evaluating software to determine whether the products of a given development phase satisfy the conditions imposed at the start of that phase, according to the requirements and design specifications (``you build the system in the
right way''); \emph{validation} is the process of evaluating software during or at the end of the development process to determine whether it satisfies specified requirements ensuring it meets the user's needs (``you build the right system'').

Within the formal methods community, verification means something much stronger, and validation something quite different, than in the IEEE 610 standard.
Loosely, within the FM community, \emph{verification} (in the paper \emph{formal verification}) means the use of formal methods to formally, statically (and usually modularly) verify that a system conforms to its specification (under all possible execution scenarios).  
On the other hand, \emph{validation} (in ther paper \emph{formal validation}) means the use of traditional software engineering practices to check, statically or dynamically, that a system conforms to its specification (whether in the form of tests, requirements, etc.) by hand or via execution under some execution scenarios.
In the present paper both perspectives are respected and acknowledged, with the attempt to be absolutely clear in using these terms unambigously.
Additionally, ``...development of a system,'' more precisely means ``...analysis, design, development, verification, formal validation of a system''.

@@

Proposing new practices with no supporting evidence is an uninteresting proposition.  
A complex case study is mandatory to test bench the proposed practices.  
Our case study focuses on the development of a device driver to control and communicate with an embedded custom circuit board equipped with more than a dozen sensors.
The communication protocol with the board is asynchronous and packet-based and the board is novel and custom-designed.  
The protocol is focused on controlling the board and reporting sensor measurements and board status.

This software driver has been developed using our process while statisfying seemingly contradictory requirements: (1)~the protocol implementation must be thoroughly verified; (2)~parts of the protocol must be formally specified; (3)~a board simulator must be designed, implemented, and verified to match the specifications of, and actual behaviour of, the physical circuit board; (4)~if board components change during development, the associated formal specifications and the simulator whose behavior reflects such changes must be updated accordingly; (5)~strict software delivery deadlines.  
Consequently, these process requirements constitute an environment that benefits from both agile and formal approaches to software development.

A novel development process has been drafted to cope with the contrasting requirements of the case study, blending agile and formal practices.  
In the following the development process and the practices are summarized, and their successful application to the case study is reported, yielding several promising results.

@@

% what is needed, high-level
To enable the integration of the two worlds, the highly iterative and incremental approach found in most of the agile methodologies must be maintained, and formal validation  methodologies, traditionally used in a waterfall development process (formally specify the system, implement the system, validate the implementation against the specification), must be adapted to a highly iterative one.

% what must be considered: TDD
The most common software verification practices in agile methodologies must be considered, when trying to integrate formal and agile methodologies, since they are a fundamental part of the development process. 
The most significant verification practice is Test Driven Development (TDD)~\cite{Beck2003}, TDD is a software development technique, originally defined in Extreme Programming (XP)~\cite{Beck2004} methodology, characterized by writing unit tests before writing the code.
Actually, the general technique, consisting in writing the tests before the code, is used at high abstraction levels as well, since user stories (the requirements) are translated into test cases (the functional tests) before implementation.
The test driven approach is a cornerstone of XP, but it is so popular that has been adopted in many other agile methodologies as well.
It is also considered a good software development technique when used on its own, regardless of the enclosing development process.

% test driven kinds
The test driven approach relies on writing the tests before the actual implementation. 
It is appliable in a very small scale (Test Driven Development) yielding to activity cycles as short as a few minutes.
It is appliable in a large scale as well (requirements immediately translated into functional tests), yeilding to activity cycles as long as a full delivery iteration (usually no longer than a couple of months).

% previous related work
Some tentative attempts to reconcile formal and agile methodologies have been developed.
The recurrent approach of the attempts consists in removing the handmade test suites, substituting them with different kinds of automated formal validation based on the system formal specification. 
The automated validation can be as simple as enabling runtime verification of assertions or automatically generating a test suite based on the system properties, or as complex as actually proving the whole system behaviour through static model and code checkers or theorem provers (formal verification).

% a quick analysis of the related works
Eleftherakis and Cowling in~\cite{Eleftherakis2003} propose XFun, an iterative and incremental process to generate X-Machines models, an extension of Finite State Machines. 
The development process impose a complete specification of the system, and its verification is done exclusively using the tools provided by the X-Machines formal methodology. 
The development process proposed in~\cite{Herranz2003b} by Herranz and Moreno-Navarro is quite similar: they propose an iterative process to model a system using SLAM-SL, an object-oriented specification language, and its tools. 
Some XP practices are fully adopted or considered compatible, as pair-programming, iterative and incremental development, system refactoring; while tests are completely replaced by the verification tools provided by the SLAM suite.
A different formal methodology, but same approach, is found also in~\cite{Suhaib2005} where XFM methodology is explained.

% previous work: a common approach
As already mentioned, in all these attempts, the handmade test suites are automatically generated using the system formal specification, or completely replaced by theorem provers or static checkers. 
The requisite to enable these approaches is, at least, a specification of the entire system under development to enable simple verification techniques as runtime verification (obtained by assertions and design by contract~\cite{Meyer1997} preconditions, postconditions and invariants) and test suites automatic generation~\cite{Cheon2002,Cheon2004,Cheon2005}.
At most, a complete and sound system specification, usually supported by an appropriate formal language (to specify model and properties) and a constrained programming language, used to actually prove properties about the system under development~\cite{CatanoHuisman02,DetlefsNelsonSaxe2005,KiniryCok04}.

% previous work: why is not satisfying
The approach is simple: the verification, documentation and requirement specification purposes of the tests are moved on to the formal specifications, and a complete formal specification of the system under development is needed to apply these methodologies.
The problem is that this requirement is hard to fulfil, since the effort required to write a complete formal specification of the system, in most real world complex cases, is usually greater than writing a suite of tests~\cite{Glass2002}.
If a complete formal specification is not built, all the previous methodologies share a common problem: the parts not formally specified are not verified at all.
These can obviously be verified with the traditional methods, but the loops between development artifacts and related activities (code, tests, design, refactoring) are neither detailed nor enforced.
The loops and relations between the different development artifacts are the inner engine of many agile methodologies, they impose an iterative and incremental pace to the development process.

% what we want to achieve
The objectives we want to achieve are the following: to formally specify only parts of the system, in order to cope with constrained resources and unstable requirements; to draft an highly iterative and incremental development process that blends formal and agile practices.

% how we want to achive it
The main specific problem, by now never addressed, is how to connect development artifacts of the two worlds together: we will define activity cycles, similar to what is found in agile methodologies (especially in XP), to solve this problem in an highly iterative development process.
We will show how tests can drive the formal specification, how the formal specifications can drive tests and code development, how handmade and automated tests can coexist and support each other, how the unspecified parts of the system can be incrementally specified.
That is, we will show how to blend, and not compose, formal validation and agile verification practices: a ``Mole'' of verification practices\footnote{A mole (\emph{'m\=ol\=a}) is a highly spiced Mexican sauce made chiefly from chili peppers and chocolate, served with meat.}.

@@

``Mole'' practices are applied in the context of a real world case study that matches the problem requirements previously detailed: unstable requirements and development artifacts that need to be formally specified and validated.

@@

% SenseTile introduction
The \textbf{UCD CASL SenseTile System} is a large-scale, general-purpose sensor system developed at the University College Dublin in Dublin, Ireland.  
The facility provides a scalable and extensible infrastructure for performing in-depth investigation into both the specific and general issues in large-scale sensor networking. 
This system integrates a sensor platform, a datastore, and a compute farm.  
The sensor platform is a custom-designed but inexpensive sensor platform (called the \ST) paired with general-purpose small-scale compute nodes, including everything from low-power PDAs to powerful touchscreen-based portable computers.  
Besides containing over a dozen sensors packaged on the \ST itself, the board is expandable as well, as new sensors are added to it easily.

% case study: the sensor board development tasks
The case study is focused in building the sensor board and its software driver concurrently.  
Because of hard time constraints and the initial unavailability of the custom board, we must progress concurrently with all the development tasks, including: (1)~specification of the communication protocol; (2)~specification of the physical sensor board; (3)~development and fabrication of the physical board; (4)~development of the embedded software for the board; (5)~development of the communication protocol software driver of the board; and (6)~development of software simulators of the board.

% what we have to do, what we outsourced, and the dependencies between tasks
The development of the physical board, along with its embedded software (development tasks~2, 3, 4), is carried out by a third party under our guidance, thus it is not directly taken into account here. 
The specification of the communication protocol (development task~1) is a joint effort between ourselves and the third party, while the remaining tasks~(5, 6) are performed in isolation.  
Dependencies are straightforward: the specification of the communication protocol~(1) is the most stable element, but it still depends on the sensor board~(2, 3, 4): large changes in the latter affect the former.
Additionally, the driver and the simulator depends directly on the communication protocol.

% communication with external manufacturer
Communication and synchronization with the external manufacturer is frequent but not optimal.  
The main synchronization artifact is the protocol specification, which remains stable at high-level, but is changed frequently in the low-level details.  
A common domain language has been found and agreed upon, but the formal methodologies and programming platforms differs widely.  
Consequently the informal specification of the communication protocol is the most reliable source of information.

@@

% hardware-software
The hardware-software formal co-design methodologies usually have several common development artifacts~\cite{Slomka2000, Hoffman2001} including a high-level specification of the system, a translation (refinement) of the high-level specification to low-level ones (hardware and software counterparts), the possibility to generate software code from the low-level specifications, a hardware simulator capable of simulating an hardware component based on hardware specification, the hardware component developed.

% device drivers
When considering the development of the device and of its software driver as separate entities that possibly have to be developed concurrently, the existing approaches are similar to the ones seen in the case of hardware-software co-design~\cite{Valderrama1995,Siegmund2002,Ryzhyk2009}, all of which focus on a specification that aims to be as complete as possible.

% complete specification is not feasible
A complete specification is neither feasible nor convenient in our case. 
The protocol specification must be enhanced incrementally and its complete specification cannot be provided.  
This makes the problem an ideal candidate to test our ``Mole'' of practices.

% formal methodology constraints
To perform this case study, an appropriate formal methodology must be chosen.  
It must be able to support a specification that is built incrementally.  
Consequently: formal methods that demand a complete specification must be avoided; additionally, to best synthesize with agile methods, the formal language must be able to automatically generate tests; finally, the formal method must be complemented by tools that support the verification of system properties both at design-time (via static analysis, for full functional verification of critical subsystems) and at runtime (for dynamic validation for integration with TDD).

\subsection{The Chosen Formal Methodology: Formal Specifications with JML}

% the chosen formal methodology
Driven by these requirements and our expertise, the Java programming language and platform are chosen for the implementation, the Java Modeling Language (JML)~\cite{LeavensBakerRuby-Prelim} is used to write formal specifications, and JML2~\cite{JML2}, and Extended Static Checker for Java (ESC/Java2)~\cite{ESCJava2,KiniryCok04SmallCite} tool suites are used as supporting tools~\cite{BurdyEtal05-STTT}.  
The JML2 tool suite includes runtime assertion checker~\cite{Cheon-Leavens02b} and a unit test generator~\cite{Cheon2002}.
ESC/Java2 supports automatic, modular, static reasoning about the correctness of an implementation against its formal specification.
These tool suites do not require a complete specification to be used effectively.

% JML introduction
JML is the de facto standard specification language for describing Java programs behaviour.  
JML is a rich behavioural interface specification language (BISL) and thus focuses on the modular specification of classes and objects in which inheritance is used solely for behavioural subtyping~\cite{LeavensBakerRuby99}.  
JML includes standard specification constructs like assumptions, assertions, axioms, and pre- and postconditions, framing clauses, and invariants.
It also includes a rich model-based specification layer that includes mathematical models, data refinements, datagroups, and many other advanced object-oriented specification features~\cite{Chalin06}.  
Many tools, ranging from compilers to static checkers to full-functional verification systems support the JML language~\cite{BurdyEtal05-STTT}.

@@

% abstraction in test suites
High level test suites, composed of functional tests, describe requirements and bind together requirement documentation and system behaviour: an informal requirement is supported by a set of functional tests. 
The associated development cycle is one iteration long. 
Low level test suites, composed of unit tests, describe the module behaviour and bind together code documentation, code and the contract imposed on the module: unit tests represent the contract and the documentation. 
The associated development cycle is less than one day (as short as a few minutes).

\subsection{``Mole'' practices}

@@

% composition of the two worlds
Both Test Driven Development (TDD) and classical Formal specification Driven Development (FDD) guarantee consistency on all the involved artifacts. 
Both TDD and FDD support iterative development.
The development cycle adopted in TDD is shown in Figure \ref{fig:tdd_diagram} and the one adopted in classic FDD is shown in Figure \ref{fig:fdd_diagram}.
Unit tests used in TDD are substituted  in FDD by formal specifications supported by the verification environment; in our case the verification environment is the generated unit test suite.
Formal specifications replace unit tests: they specify, document and verify the system under development.

\begin{figure}[ht]
  \begin{minipage}[b]{0.48\linewidth}
    \centering
    \includegraphics[scale=0.5]{diagrams/TDD_classic}
    \caption{TDD development cycle.}
    \label{fig:tdd_diagram}
  \end{minipage}
  \hspace{0.02\linewidth}
  \begin{minipage}[b]{0.48\linewidth}
    \centering
    \includegraphics[scale=0.5]{diagrams/FDD_classic}
    \caption{FDD classic development cycle.}
    \label{fig:fdd_diagram}
  \end{minipage}
\end{figure}

% composition is not enough
Is this enough to deal with formal and informal artifacts?
FDD is used for the formal specified parts of the system, whether TDD is used for the remaining ones: a ``composition'' of formal and agile verification practices.
Yet, it is not enough.
There are several problems that still need to be addressed.
No communication between the two worlds is permitted, it is not defined how to formally specify a module after it has been implemented and verified in a non formal way.
A module only partially specified is a sort of hybrid that must be treated accordingly.
Finally, in the FDD cycle shown in Figure \ref{fig:fdd_diagram}, the implicit assumption is that it is always possible to identify the correct formal specifications of the system, but this is not always straighforward.
Identify the correct formal specification is a difficult task, it has to be supported and verified by the development process.

% FDD with handmade tests
A partially specified module cannot be verified properly only through automatically generated tests.
Handmade tests have to be used together with generated ones.
Handmade tests verify complex behaviours that are not formally specified, or complex behaviours that are formally specified but cannot be replicated by the formal verification tools, because of the limits of the verification tools themselves. 
The constraints imposed by formal verification tools~\cite{Glass2002} must be taken into account when deciding whether supporting the code with handmade tests, and verification tools based on automatically generated tests are not an exception.
The development cycle able to blend formal specifications, handmade tests and code development is shown in Figure \ref{fig:fdd_hand_made_tests_diagram}: the formal specification drives the handmade tests. 
Both handmade tests and the formal specification drive the code development.

\begin{figure}[ht]
  \begin{minipage}[b]{0.48\linewidth}
    \centering
    \includegraphics[scale=0.5]{diagrams/FDD_hand_made_tests}
    \caption{FDD development cycle with handmade tests.}
    \label{fig:fdd_hand_made_tests_diagram}
  \end{minipage}
  \hspace{0.02\linewidth}
  \begin{minipage}[b]{0.48\linewidth}
    \centering
    \includegraphics[scale=0.5]{diagrams/TDF}
    \caption{TDF development cycle.}
    \label{fig:tdf_diagram}
  \end{minipage}
\end{figure}
